---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggdebug)
})
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.height = 2.5
)

set.seed(1)

# mtcars <- mtcars %>% mutate(cyl = as.factor(cyl))
```
# ggdebug

<!-- badges: start -->
![](https://img.shields.io/badge/lifecycle-alpha-orange.svg)
<!-- badges: end -->

`ggdebug` is a package for debugging ggplot2 stats.

I created this package to help in developing new `ggplot2` stats, and trying
to decipher and understand how existing ones work.

What's in the box:

* `create_stat_with_caching()`
    * **Capture arguments and the return values internal to Stat methods**
* `inject_data_into_stat_function()`
    * **Inject arguments and return values directly into Stat methods** 


This package wouldn't be possible without:

* [ggplot2](https://cran.r-project.org/package=ggplot2) by Hadley Wickham
* [Brodie Gaslam's](https://twitter.com/BrodieGaslam) great guide to [ggplot2 internals](https://htmlpreview.github.io/?https://github.com/brodieG/ggbg/blob/development/inst/doc/extensions.html#stat-compute)
* [boBRudis's](https://twitter.com/hrbrmstr) post on [demystifying ggplot2](https://rud.is/books/creating-ggplot2-extensions/demystifying-ggplot2.html)


## Installation

You can install the development version from [GitHub](https://github.com/coolbutuseless/ggdebug) with:

``` r
# install.packages("remotes")
remotes::install_github("coolbutuseless/ggdebug")
```


## Debug `geom_histogram` + `StatBin`

The following example shows how `ggdebug` can be used to capture the data during 
creation of `geom_histogram()`.   

`geom_histogram()` uses a default Stat of `StatBin`.  Instead we create a 
debugging version of this Stat and use it explicitly.

The captured data is available by calling
`ggdebug::get_cache_data()`.


```{r setup, fig.height = 3}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create a new stat which is identical to StatBin, but captures the
# data for all the methods.
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
StatBinDebug <- ggdebug::create_stat_with_caching(
  StatBin
)
```


```{r example1, collapse = TRUE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create the plot using the new Stat
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
ggplot(mtcars) +
  geom_histogram(aes(mpg, fill = as.factor(cyl)), stat = StatBinDebug, bins = 10) +
  theme_bw() +
  facet_wrap(~cyl)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Get the data we captured from this process
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
cdata <- ggdebug::get_data_cache()

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Interrogate the captured data. 
# Each method has its only entry in the data cache, and 
# within each entry there are both the `args` and the `return` value
# for this method
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
names(cdata)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# arguments and the return value for the 'parameters()' method
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
cdata$parameters

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# The 'data' argument passed in to the 'compute_layer()' method
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
head(cdata$compute_layer$args$data)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# The return value from the 'finish_layer()' method
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
head(cdata$finish_layer$return)
```


## Inject data into `geom_histogram` + `StatBin`

`ggdebug` also allows you to inject arguments and override return values within 
methods inside a Stat.

In this example, we take the above histogram, and at the last stage (`finish_layer`)
we inject a modified return value to override the real one.




```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# From the above debugging example, grab the return value from `finish_layer`
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
finish_layer_return_value <- cdata$finish_layer$return

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Change the visual properties for the bar with the greatest height
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
finish_layer_return_value <- finish_layer_return_value %>% 
  group_by(group) %>%
  mutate(
    fill   = ifelse(count == max(count), '#000000', fill  ),
    colour = ifelse(count == max(count), '#ff0000', colour),
    size   = ifelse(count == max(count), 1        , size  )
  ) %>%
  ungroup()

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create a modified version of StatBin which only returns our version of the 
# dataset from `finishlayer`
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
StatBinInject <- inject_data_into_stat_function(
  StatBin, 'finish_layer',
  return_value = finish_layer_return_value
)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Recreate the plot using the injected data as the return from `finish_layer`
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
ggplot(mtcars) +
  geom_histogram(aes(mpg, fill = as.factor(cyl)), stat = StatBinInject, bins = 5) +
  theme_bw() +
  facet_wrap(~cyl) +
  labs(title = "Use injected data to highlight the maximum bar within each group")
```





## Inject tetris pieces into `geom_histogram` + `StatBin`

```{r}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create 3 tetris pieces
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
tetris <- tribble(
  ~fill    , ~xmin, ~xmax, ~ymin, ~ymax,
  '#ff0000', 10, 19, 6, 7,
  '#ff0000', 13, 16, 6, 8,
  '#00ff00', 25, 28, 6, 8,
  '#00ff00', 28, 31, 7, 9,
  '#0000ff', 25, 34, 6, 7,
  '#0000ff', 28, 31, 6, 8
)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Assign each piece to a different panel
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
tetris <- tetris %>% mutate(
  PANEL    = factor(c(1, 1, 2, 2, 3, 3), levels = 1:3), 
  group    = 1,
  linetype = 1,
  x        = xmax,
  y        = ymax
)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Add the tetris data to the original data returned by `finish_layer`
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
finish_layer_return_value <- bind_rows(
  tetris, 
  cdata$finish_layer$return
)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create a StatBin that only returns this modified data from `finish_layer`
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
StatBinInject <- inject_data_into_stat_function(
  StatBin, 'finish_layer',
  return_value = finish_layer_return_value
)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create the plot to generate tetris pieces injected into the data
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
ggplot(mtcars) +
  geom_histogram(aes(mpg, fill = as.factor(cyl)), stat = StatBinInject, bins = 5) +
  theme_bw() +
  facet_wrap(~cyl) +
  labs(title = "Use injected data to inject tetris pieces")
```











